\chapter{Project Plan}

\section{Aims}
To learn about the family of Paxos-like consensus protocols and then to formulate a proof of their correctness by implementing a library of reusable verified distributed components using Disel. Disel is a framework for compositional verification of distributed protocols, built on top of Coq proof assistant, to verify correctness of the implemented components.
Hence, this project will help me learn about the workings of distributed protocols and how to reason about their correctness.

\section{Objectives}
\begin{enumerate}
  \item Read about and understand the classical Paxos-like consensus algorithms. Develop state transition systems for the algorithms and identify the invariants that need to be preserved during the operation of the algorithm.
  \item Implement a simulation of the protocols in Python.
  \item Formulate the implemented protocols in Disel by using the developed state-transition systems.
  \item Mechanise the proofs of the identified protocol invariants in Disel/Coq.
  \item Add additional communication channels and prove composite invariants.
  \item Provide an abstract specification of the protocol, usable by third-party clients.
  \item Mechanise a client application of the protocol verified out of the abstract interface.
\end{enumerate}

\section{Expected Deliverables}
\begin{itemize}
  \item Descriptions of the state-transition system and invariants of the Paxos-like Consensus protocols.
  \item Python implementation of simulations of the Paxos-like Consensus protocols.
  \item Proofs of correctness of the implemented protocols in Disel. Thereby, providing a library of reusable verified distributed components written in Disel/Coq.
\end{itemize}

\section{Work Plan}
\begin{itemize}
\itemsep0em
  \item Project start to end October (4 weeks).
    \begin{itemize}
      \item Read about and understand the workings of Paxos-like Consensus protocols.
      \item Implement the protocols in Python.
    \end{itemize}
  \item November start to mid-December (6 weeks).
    \begin{itemize}
      \item Create state transition diagrams and identify invariants of the protocols.
      \item Formulate the protocols in Disel.
      \item Prove correctness of the identified protocol invariants in Disel/Coq.
    \end{itemize}
  \item Mid-December (8 weeks) to mid February.
    \begin{itemize}
      \item Iterate on the Disel proof to make it more concise.
      \item Formulate the abstract interface.
      \item Verify a simple client application.
    \end{itemize}
  \item January start to March end (12 weeks).
    \begin{itemize}
      \item Submit Interim Report (due in late January).
      \item Work on Final Report.
    \end{itemize}
\end{itemize}
